The Talking Heads Experiment

### 2015-04-10 chap2/chap2.tex
fixes
cumulative changes to preface and chapter 1
updated labels, image widths, quotation marks, capitalisation, etc. for chapters 1-7

%\documentclass{../_combined/fcg-book}
\chapter{Preview}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap1/figs/teleportation}}
\caption{ The Talking Heads agents are implemented as software entities that can travel over the internet. 
To play a game, they get downloaded in a local server that drives the cameras and orchestrates a game. After a game, the software 
state is uploaded again to travel towards another location.}
\label{f:teleport}
\end{figure}

However, in the case of computer-based \is{teleportation}
artificial agents, we {\itshape can} 
make the distinction. It is possible to capture the mental state
of an agent in software, load this into a physical 
body, and then operate 
the agent. Afterwards the agent can extract himself
again from the body, teleport himself 
to another physical location through a data transmission network
like the Internet, get instantiated there in another
body, and experience another reality and physically meet
other agents. This is exactly how we have implemented the 
Talking Heads experiment.\footnote{
The agent teleportation infrastructure is in itself a 
fascinating non-trivial engineering project. Contributions 
from Angus McIntyre, Alexis Agahi, Sylvere Tajan and
Frederic Kaplan are gratefully acknowledged (\cite{McIntyre:1999}). 
The fact that agents can teleport proves that we are dealing 
with a truly distributed multi-agent system. It also 
introduces physical parallelism in the agent-agent and 
agent-environment interactions. 
For a general introduction into multi-agent system
technologies and design methodologies, 
see \cite{Ferber:1999}.}
There are on the one hand
the physical structures, which I will refer to as the {\it
robot bodies}. They are installed in different physical 
locations somewhere in the world and connected
with each other through the Internet. Then there is 
a population of software structures that 
are occasionally loaded and instantiated in specific
robots. I will call these software structures
{\itshape virtual agents}. 
A {\itshape real agent} (a Talking Head) 
only exists when the virtual agent is
loaded in a physical robot body.

Virtual agents cannot interact and 
an interaction between two `real' agents
can only take place when they are both physically present
in the same location. Thus an agent can travel from
Paris to Tokyo at the blink of an eye, rather than having
to take a plane, but two agents can only interact when they
are instantiated in the same physical environment. 
It is in principle possible that agents in different 
physical locations describe to each other the environment that they 
see (but which the other one does not see), just as we 
would do in a telephone conversation. This is only possible,
however, after the agents have had sufficient interactions
with each other in a shared physical world to have 
developed and learned a grounded shared language.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap1/figs/Interface}}
\caption{ Interface through which the internal states of agents can 
be inspected. Two agents are shown. The top windows show the state of an agent, 
the middle windows the camera inputs that the agent sees and the bottom windows show 
their discrimination trees.}
\label{f:agentview}
\end{figure}

In constructing the robot bodies, we have used as much as
possible off-the-shelf standard 
components so that we could focus almost completely on issues
directly relevant to language and meaning.
Each robot's low-level vision system
(integrated in the camera) is already very sophisticated.\footnote{
The camera is a Sony EVI-D31. The main computer is a Power 
Macintosh from Apple, Inc. The agent servers run under the 
Linux operating system.}
It can focus automatically to get a sharper image
and autonomously track a moving object. The speech signal is 
produced with a standard text-to-speech system 
so that we did not have to 
worry about building complex audio modules ourselves.
The computer hardware 
is powerful, but not specialised nor in the supercomputer 
range. All programs have been written in the standard symbolic 
programming language of artificial intelligence 
research, namely LISP.\footnote{
The construction of the programs underlying the 
Talking Heads experiment has required advanced
artificial intelligence programming techniques, 
such as discussed by \cite{Norvig:1996}. A general
toolkit for the systematic execution of 
simulation and physical experiments, called BABEL, has been
designed and implemented by Angus McIntyre. The 
toolkit allows the definition and modular composition
of cognitive architectures, the design of 
experiments, and the monitoring and displaying of 
results, see \cite{McIntyre:1998}.} What makes the Talking Heads experiment
special is not the hardware or software tools
but what we have done with it.

@@

Agents can only engage in interactions with other agents 
when they are physically instantiated
in a robot. Each agent has a basic brain architecture
with different layers performing the cognitive functions 
relevant for playing language games: 
\begin{itemize}
\item A perceptual layer which performs low-level signal 
processing to segment the image and collect data about 
each segment such as the colour, size, position or shape
of a segment. 
\item A conceptual layer which categorises and 
conceptualises the segmented and processed image. 
It is based on a self-generated and evolving repertoire
of categorial distinctions, such as red versus green, or
small versus large. Such a repertoire is referred to as 
the agent's ontology in this book. 
\item A lexical layer which maintains an evolving repertoire
of associations between meanings and words, which I will refer to 
as the lexicon, and performs lexical lookup while parsing 
or producing utterances. 
\item A syntactic layer which uses 
grammatical schemata for organising words
in larger structures or for recognising these structures and
reconstructing complex meanings. 
\item A pragmatic layer which carries out the
scripts for playing language games and maintains the machinery 
for engaging in interactions with other agents in a 
shared environment. 
\end{itemize}
Each of these layers is described in more detail 
later. The internals of the layers are not static but
constantly evolving and adapting. They are not strictly 
modular but coupled in various ways to each other.
Each verbal interaction 
in effect changes the agent's internal state and thus 
influences future behaviour. A new, virgin agent 
starts without any built-in ontology, lexicon, or grammar. 
This is one of the crucial points of the whole experiment, 
because we want to test possible theories on how language
and meaning can evolve and be acquired {\itshape ab initio}.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap2/figs/interface}}
\caption{ Internet interface through which users can 
access the state of games on remote sites and 
follow the experiment.}
\label{f:plate8}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap1/figs/Whiteboard}}
\caption{ The physical environment of the Talking Heads  
consists of a white board on which various
geometric figures can be pasted. The light conditions were not under the control 
of the experimenters in most locations.}
\label{f:plate9}
\end{figure}

@@

The speaker first looks at one area of the white board and 
directs the attention of the hearer to the same area.\footnote{
I will explain later how exactly agents
decide on a particular scene and how they are able to 
draw each other's attention to specific areas of the 
white board.}
The objects located in this area constitute the {\itshape context}. 
The speaker then chooses one object from the context, which 
I will call the {\itshape topic}, and gives a verbal hint
to the hearer. The {\itshape verbal hint} is an expression
that identifies the topic 
with respect to the other objects in the context. For example, 
if the context contains a red square, a blue triangle, 
and a green circle, then the speaker may say something like 
`the red one' to identify the red square as the topic. 
If the context contains also a red triangle, he has to be more 
precise and say something like `the red square' to delineate
the topic from the red triangle as well as
from the blue square. 
Of course, the Talking Heads do not say ``the red square'
but use their own language and concepts which are not 
necessarily the same as those used in English. For example, 
they may say ``malewina' to mean [UPPER EXTREME-LEFT LOW-REDNESS].

Based on the verbal hint, the hearer tries to guess what
topic the speaker has chosen, and communicates his choice 
to the speaker by pointing to the object. Given that 
the robots do not have arms, pointing is realised by 
focusing on an object. One robot can `see' in which direction another one 
is looking, and thus know where he is `pointing'. 
The game succeeds if the topic guessed by the hearer is 
equal to the topic chosen by the speaker. 
The game fails if the guess was wrong or 
if the speaker or the hearer failed at some earlier point in the 
game.

@@

The environment of the Talking Heads is 
not fixed. The human experimenter 
may change the position of objects, add new kinds of 
objects, or eliminate
others. Consequently a strategy of naming individual objects
will not work. It would lead to a proliferation of 
proper names and it would require the Talking Heads to
recognise objects, which is very difficult to do.\footnote{
For a thorough exposition of the difficulties of 
object recognition, see \cite{Ullman:1996}. 
Object constancy comes fairly late in the 
acquisition of a child's ontology, as Piaget's conservation
experiments have shown. Language probably plays an important
role in forming the notion of an object.}
Indeed, humans don't exclusively use proper names in 
natural language conversations 
either. We say `could you give me the red small square' as 
opposed to `could you give me O_143'. Natural language 
words like `red' or `small' name perceptually grounded categories
and syntactic structures indicate how they should
be combined and used to find the topic. The relation between 
a language expression and its referent is therefore 
always indirect. This is summarised
in the semiotic square (\figref{triangle}), which I will use throughout this book to 
help understand and analyse the nature of language communication. \is{semiotic square}
The semiotic square relates the four entities
involved in a verbal interaction: 
\begin{itemize}
\item An {\itshape utterance}, such as `small red square', which is 
transmitted as a physical 
signal from one agent to another one through the 
external environment. It is written between double quotes. 
\item A {\itshape meaning}, which consists of categories like [RED], [SMALL], 
or combinations of categories, like \{[RED] [SMALL]\}. Labels of 
categories are written in capital letters between 
square brackets. 
\item An {\itshape image segment}, denoted as $S143$, which is 
a segment of the image perceived through the camera.
\item A {\itshape referent}, like object $O143$, which is 
an entity in the real world. 
\end{itemize}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap2/figs/triangle}}
\caption{\label{triangle} Any verbal interaction
involves four entities here grouped in the
{\it semiotic square}. The relation between utterance
and referent always needs to be established indirectly by 
passing through perception and meaning.}
\end{figure}

The systematic relation between meaning and referent is 
usually studied under the heading of semantics, and the 
systematic relation between meaning and utterance as
grammar (including syntax, morphology and lexicon).\footnote{
For a general introduction to the contemporary linguistic
viewpoint on the processes involved, see: \cite{Vanvalin:1997}. 
In a logical approach to language, as 
exemplified by Montague grammar (\cite{Montague:1974}),  
meanings are represented using a logical formalism, i.e.
a variant of intensional logic. \is{Montague grammar}
Natural language expressions are systematically 
related to expressions in this logic, and a
formal semantics system defines how expressions in the logic
are mapped onto their denotations. Because this is 
a formal framework, the denotations consist of formal 
models. To make the Talking Heads experiment work, 
we needed to develop a grounded semantics system, which details
how an agent may go from physical reality to meaning
using a perceptual apparatus, and from meaning 
to physical reality. The logical structure of 
the meanings we will investigate are very simple 
(unary predicates and conjunctions). But once we know how to 
set up a grounded semantics for simple meaning
structures we can scale it up to the more complex meaning
structures typically studied in logic.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.85\textwidth]{chap2/figs/triangle2}}
\caption{\label{triangle2a} Left: processes carried
out by the speaker. Right: processes carried out by the hearer.
There are also feedback processes moving in alternate directions
until the agents settle on coherent choices for all the items
in their semiotic squares.}
\end{figure}

@@

Each of these activities requires knowledge and/or skill
(summarised in the table below). Perceiving requires 
visual processes capable of segmenting images and deriving image 
segments. Conceptualisation requires an ontology, 
a repertoire of perceptually grounded distinctions
that can be applied to a segmented image to yield 
distinctive categories or category combinations that 
may constitute the meaning of the utterance. Verbalising 
a conceptualisation requires 
a {\itshape lexicon} that maps parts of the meaning to words and 
a {\itshape syntax} that specifies how to organise individual words
into a larger complex. 
The hearer must use similar 
knowledge sources in the other direction. He must use 
his lexicon and syntax to reconstruct the meanings expressed
by the utterance, and then use the 
ontology again to apply the meaning to the present 
context to find the referent \tabref{tab:2:findthereferent}.

@@

I deliberately left an important aspect of language out. 
In the case of physical agents, the form cannot be transmitted
directly but needs to be articulated in speech sounds, 
written signs, or gestures, to create a true utterance. This additional 
complexity will not be discussed further in this book - even though 
it is a fascinating topic in its own right.\footnote{
See for a state of the art review: \cite{Clark:1990}. 
We have already been conducting quite extensive 
research in our group on the origin of sound systems 
using similar principles as the ones discussed in this 
book for the origins of word semantics. 
See: \cite{DeBoer:1997}. The complex 
adaptive systems approach underlying this phonetics
work was already foreshadowed in work of phoneticians
like \cite{Liljencrants:1972}.}
To make a verbal interaction nevertheless complete, agents
are given a repertoire of consonants and vowels with 
which they can make random syllables and syllable combinations, 
like `wabido', `bimaku', etc. The articulation and 
recognition of these syllables is assumed to be acquired
already and transmission is engineered to be error-free. 
This way, our attention can be focused on how
ontologies, lexicons, and grammars may emerge.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap2/figs/hpos}}
\caption{ Three examples of segmented images (from top to bottom). Left and right images 
show the perceived and segmented images for two agents respectively. These images are always
different because they are dependent on the position of the agent. 
The topic is indicated by a dashed bounding box in the 
image of the speaker (on the right for the first case and on the left for the others). 
Segments which are too small are ignored. The topics have all been conceptualised
as being to the right and so the same word 
`gofubo' has been used to successfully refer to them.}
\label{f:plate10}
\end{figure}

Segmentation can happen according to several criteria. \is{image segmentation}
For example, patches with similar colour can be grouped 
into a single patch, edges can be identified 
and then linked with each other to form line segments
circumscribing the contours of 
an object, or consecutive images can be compared to extract the parts
that changed and hence moved as a single object. 
There is now a solid body of techniques from decades
of machine vision research to efficiently segment scenes according to 
these and other methods. The Talking Heads use several methods
in parallel and combine their output to get a clearly segmented
picture.\footnote{
For general introductions to these areas, 
see: \cite{Ballard:1982} and \cite{Fischler:1987}}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap2/figs/scene1-1}}
\caption{\label{scene1-1} The scene contains two 
objects: a triangular object and a rectangular object
with the same gray levels. 
Each one is characterised by values on three sensory 
channels: HPOS, VPOS, and GRAY.}
\end{figure}

The low-level visual processes outputting
values on the various sensory channels are already quite
complicated in themselves. I will discuss them further in 
Chapter 3. I will then argue that the 
agent's repertoire of visual processes does not have to be
static and programmed in advance but evolves and adapts
through a selectionist process. New processes may `grow'
by the random combination of primitive operations and are
pruned when they fail to yield useful information or information 
which is irrelevant in the environment in which the agent 
finds himself.

@@

The data on sensory channels are values from a continuous domain
(between 0.0 and 1.0). 
To be the basis of natural language communication, these
values must be transformed into a discrete domain. This is 
precisely the task of the conceptual layer. It 
performs massive data reduction to make infinitely
rich environments managable. One means of categorisation
is to divide up each domain of values on a particular
sensory channel into regions and 
simply assign a category to each region, thus creating a 
discrimination tree.\is{discrimination tree} For example, 
the HPOS-channel can be cut in two halves leading to a 
distinction between [LEFT] ($0.0 \leq HPOS < 0.5$) and [RIGHT] 
($0.5 \leq HPOS \leq 1.0$). The triangular object 
in \figref{scene1-1} has the value HPOS=0.35 
and would therefore be categorized as [LEFT]. Similarly, the 
VPOS-channel can be divided in two halves yielding the 
categories [LOWER] and [UPPER], and likewise the GRAY-channel
yielding the categories [LIGHT] and [DARK]. 
Given these categories, 
the rectangular object in the scene of 
\figref{scene1-1} would be categorized as [RIGHT UPPER LIGHT].
Of course, light, dark, left, lower, etc. are labels 
that I have given to these categories. The Talking Heads 
create categories by partitioning sensory channels but do
not use these labels internally.

It is always possible to refine a distinction by further
subdividing its region. Thus an agent could further divide the 
bottom region of the HPOS-channel (categorized as 
[LEFT]) into two subregions [TOTALLY-LEFT] ($0.0 \le HPOS < 0.25$), 
and [MID-LEFT] ($0.25 \le HPOS < 0.5$). The triangular object
can now be categorized as [MID-LEFT], if simply categorising it
as [LEFT] is not distinctive enough. Each of these categories can 
still be further refined. The categorisation networks
resulting from these consecutive binary subdivisions form
discrimination trees (\figref{tree1}). It is not at
all assumed that all agents have the same trees; due to 
different developmental histories, divergence is inevitable.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.35\textwidth]{chap2/figs/tree1}}
\caption{\label{tree1} A discrimination tree displays
the divisions of the total range of 
values on a sensory channel into finer and finer subregions. 
Categories are assigned to each region at different 
levels of the tree.}
\end{figure}

There are obviously other ways to 
move from the continuous domain of sensory channels to 
the discrete domain of categories.\footnote{
See: \cite{Taylor:1989}}
For example, it is not really necessary 
to have a binary division, we could just as well split 
each region into three or more subregions. Or we could introduce
focal (prototypical) values and associate
a category with each of them. In the latter case, 
the categorisation process consists 
in identifying the prototype that is closest to 
an object's value. For the current experiment, I will stick however
to a binary categorisation strategy because that is
the simplest to understand and formally investigate.

>>>>

%\documentclass{../_combined/fcg-book}
\chapter{Preview}
\label{chap:2}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap1/figs/teleportation.pdf}}
\caption{The Talking Heads agents are implemented as software entities that can travel over the internet. 
To play a game, they get downloaded in a local server that drives the cameras and orchestrates a game. After a game, the software 
state is uploaded again to travel towards another location.}
\label{f:teleport}
\end{figure}

However, in the case of computer-based \is{teleportation}
artificial agents, we {\itshape can} 
make the distinction. It is possible to capture the mental state
of an agent in software, load this into a physical 
body, and then operate 
the agent. Afterwards the agent can extract himself
again from the body, teleport himself 
to another physical location through a data transmission network
like the Internet, get instantiated there in another
body, and experience another reality and physically meet
other agents. This is exactly how we have implemented the 
Talking Heads experiment.\footnote{
The agent teleportation infrastructure is in itself a 
fascinating non-trivial engineering project. Contributions 
from Angus McIntyre, Alexis Agahi, Sylvere Tajan and
Frederic Kaplan are gratefully acknowledged (\citealt{McIntyre:1999}). 
The fact that agents can teleport proves that we are dealing 
with a truly distributed multi-agent system. It also 
introduces physical parallelism in the agent-agent and 
agent-environment interactions. 
For a general introduction into multi-agent system
technologies and design methodologies, 
see \cite{Ferber:1999}.}
There are on the one hand
the physical structures, which I will refer to as the {\it
robot bodies}. They are installed in different physical 
locations somewhere in the world and connected
with each other through the Internet. Then there is 
a population of software structures that 
are occasionally loaded and instantiated in specific
robots. I will call these software structures
{\itshape virtual agents}. 
A {\itshape real agent} (a Talking Head) 
only exists when the virtual agent is
loaded in a physical robot body.

Virtual agents cannot interact and 
an interaction between two ``real'' agents
can only take place when they are both physically present
in the same location. Thus an agent can travel from
Paris to Tokyo at the blink of an eye, rather than having
to take a plane, but two agents can only interact when they
are instantiated in the same physical environment. 
It is in principle possible that agents in different 
physical locations describe to each other the environment that they 
see (but which the other one does not see), just as we 
would do in a telephone conversation. This is only possible,
however, after the agents have had sufficient interactions
with each other in a shared physical world to have 
developed and learned a grounded shared language.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap1/figs/Interface.pdf}}
\caption{Interface through which the internal states of agents can 
be inspected. Two agents are shown. The top windows show the state of an agent, 
the middle windows the camera inputs that the agent sees and the bottom windows show 
their discrimination trees.}
\label{f:agentview}
\end{figure}

In constructing the robot bodies, we have used as much as
possible off-the-shelf standard 
components so that we could focus almost completely on issues
directly relevant to language and meaning.
Each robot's low-level vision system
(integrated in the camera) is already very sophisticated.\footnote{
The camera is a Sony EVI-D31. The main computer is a Power 
Macintosh from Apple, Inc. The agent servers run under the 
Linux operating system.}
It can focus automatically to get a sharper image
and autonomously track a moving object. The speech signal is 
produced with a standard text-to-speech system 
so that we did not have to 
worry about building complex audio modules ourselves.
The computer hardware 
is powerful, but not specialised nor in the supercomputer 
range. All programs have been written in the standard symbolic 
programming language of artificial intelligence 
research, namely \textsc{lisp}.\footnote{
The construction of the programs underlying the 
Talking Heads experiment has required advanced
artificial intelligence programming techniques, 
such as discussed by \cite{Norvig:1996}. A general
toolkit for the systematic execution of 
simulation and physical experiments, called \textsc{babel}, has been
designed and implemented by Angus McIntyre. The 
toolkit allows the definition and modular composition
of cognitive architectures, the design of 
experiments, and the monitoring and displaying of 
results, see \cite{McIntyre:1998}.} What makes the Talking Heads experiment
special is not the hardware or software tools
but what we have done with it.

@@

Agents can only engage in interactions with other agents 
when they are physically instantiated
in a robot. Each agent has a basic brain architecture
with different layers performing the cognitive functions 
relevant for playing language games: 
\begin{itemize}
\item A perceptual layer which performs low-level signal 
processing to segment the image and collect data about 
each segment such as the colour, size, position or shape
of a segment. 
\item A conceptual layer which categorises and 
conceptualises the segmented and processed image. 
It is based on a self-generated and evolving repertoire
of categorial distinctions, such as red versus green, or
small versus large. Such a repertoire is referred to as 
the agent's ontology in this book. 
\item A lexical layer which maintains an evolving repertoire
of associations between meanings and words, which I will refer to 
as the lexicon, and performs lexical lookup while parsing 
or producing utterances. 
\item A syntactic layer which uses 
grammatical schemata for organising words
in larger structures or for recognising these structures and
reconstructing complex meanings. 
\item A pragmatic layer which carries out the
scripts for playing language games and maintains the machinery 
for engaging in interactions with other agents in a 
shared environment. 
\end{itemize}
Each of these layers is described in more detail 
later. The internals of the layers are not static but
constantly evolving and adapting. They are not strictly 
modular but coupled in various ways to each other.
Each verbal interaction 
in effect changes the agent's internal state and thus 
influences future behaviour. A new, virgin agent 
starts without any built-in ontology, lexicon, or grammar. 
This is one of the crucial points of the whole experiment, 
because we want to test possible theories on how language
and meaning can evolve and be acquired ab initio\todo{GSR: no italics in commonly used loan words}.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap2/figs/interface.pdf}}
\caption{Internet interface through which users can 
access the state of games on remote sites and 
follow the experiment.}
\label{f:plate8}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap1/figs/Whiteboard.pdf}}
\caption{The physical environment of the Talking Heads  
consists of a white board on which various
geometric figures can be pasted. The light conditions were not under the control 
of the experimenters in most locations.}
\label{f:plate9}
\end{figure}

@@

The speaker first looks at one area of the white board and 
directs the attention of the hearer to the same area.\footnote{
I will explain later how exactly agents
decide on a particular scene and how they are able to 
draw each other's attention to specific areas of the 
white board.}
The objects located in this area constitute the {\itshape context}. 
The speaker then chooses one object from the context, which 
I will call the {\itshape topic}, and gives a verbal hint
to the hearer. The {\itshape verbal hint} is an expression
that identifies the topic 
with respect to the other objects in the context. For example, 
if the context contains a red square, a blue triangle, 
and a green circle, then the speaker may say something like 
`the red one' to identify the red square as the topic. 
If the context contains also a red triangle, he has to be more 
precise and say something like `the red square' to delineate
the topic from the red triangle as well as
from the blue square. 
Of course, the Talking Heads do not say `the red square'
but use their own language and concepts which are not 
necessarily the same as those used in English. For example, 
they may say `malewina' to mean [UPPER EXTREME-LEFT LOW-REDNESS].

Based on the verbal hint, the hearer tries to guess what
topic the speaker has chosen, and communicates his choice 
to the speaker by pointing to the object. Given that 
the robots do not have arms, pointing is realised by 
focusing on an object. One robot can ``see'' in which direction another one 
is looking, and thus know where he is ``pointing''. 
The game succeeds if the topic guessed by the hearer is 
equal to the topic chosen by the speaker. 
The game fails if the guess was wrong or 
if the speaker or the hearer failed at some earlier point in the 
game.

@@

The environment of the Talking Heads is 
not fixed. The human experimenter 
may change the position of objects, add new kinds of 
objects, or eliminate
others. Consequently a strategy of naming individual objects
will not work. It would lead to a proliferation of 
proper names and it would require the Talking Heads to
recognise objects, which is very difficult to do.\footnote{
For a thorough exposition of the difficulties of 
object recognition, see \cite{Ullman:1996}. 
Object constancy comes fairly late in the 
acquisition of a child's ontology, as Piaget's conservation
experiments have shown. Language probably plays an important
role in forming the notion of an object.}
Indeed, humans don't exclusively use proper names in 
natural language conversations 
either. We say `could you give me the red small square' as 
opposed to `could you give me O\_143'. Natural language 
words like `red' or `small' name perceptually grounded categories
and syntactic structures indicate how they should
be combined and used to find the topic. The relation between 
a language expression and its referent is therefore 
always indirect. This is summarised
in the semiotic square (\figref{triangle}), which I will use throughout this book to 
help understand and analyse the nature of language communication. \is{semiotic square}
The semiotic square relates the four entities
involved in a verbal interaction: 
\begin{itemize}
\item An {\itshape utterance}, such as `small red square', which is 
transmitted as a physical 
signal from one agent to another one through the 
external environment. It is written between double quotes. 
\item A {\itshape meaning}, which consists of categories like [RED], [SMALL], 
or combinations of categories, like \{[RED] [SMALL]\}. Labels of 
categories are written in capital letters between 
square brackets. 
\item An {\itshape image segment}, denoted as $S143$, which is 
a segment of the image perceived through the camera.
\item A {\itshape referent}, like object $O143$, which is 
an entity in the real world. 
\end{itemize}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap2/figs/triangle.pdf}}
\caption{\label{triangle}Any verbal interaction
involves four entities here grouped in the
{\it semiotic square}. The relation between utterance
and referent always needs to be established indirectly by 
passing through perception and meaning.}
\end{figure}

The systematic relation between meaning and referent is 
usually studied under the heading of semantics, and the 
systematic relation between meaning and utterance as
grammar (including syntax, morphology and lexicon).\footnote{
For a general introduction to the contemporary linguistic
viewpoint on the processes involved, see: \cite{Vanvalin:1997}. 
In a logical approach to language, as 
exemplified by Montague grammar (\citealt{Montague:1974}),  
meanings are represented using a logical formalism, i.e.
a variant of intensional logic. \is{Montague grammar}
Natural language expressions are systematically 
related to expressions in this logic, and a
formal semantics system defines how expressions in the logic
are mapped onto their denotations. Because this is 
a formal framework, the denotations consist of formal 
models. To make the Talking Heads experiment work, 
we needed to develop a grounded semantics system, which details
how an agent may go from physical reality to meaning
using a perceptual apparatus, and from meaning 
to physical reality. The logical structure of 
the meanings we will investigate are very simple 
(unary predicates and conjunctions). But once we know how to 
set up a grounded semantics for simple meaning
structures we can scale it up to the more complex meaning
structures typically studied in logic.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.85\textwidth]{chap2/figs/triangle2.pdf}}
\caption{\label{triangle2a}Left: processes carried
out by the speaker. Right: processes carried out by the hearer.
There are also feedback processes moving in alternate directions
until the agents settle on coherent choices for all the items
in their semiotic squares.}
\end{figure}

@@

Each of these activities requires knowledge and/or skill
(summarised in the table below). Perceiving requires 
visual processes capable of segmenting images and deriving image 
segments. Conceptualisation requires an ontology, 
a repertoire of perceptually grounded distinctions
that can be applied to a segmented image to yield 
distinctive categories or category combinations that 
may constitute the meaning of the utterance. Verbalising 
a conceptualisation requires 
a {\itshape lexicon} that maps parts of the meaning to words and 
a {\itshape syntax} that specifies how to organise individual words
into a larger complex. 
The hearer must use similar 
knowledge sources in the other direction. He must use 
his lexicon and syntax to reconstruct the meanings expressed
by the utterance, and then use the 
ontology again to apply the meaning to the present 
context to find the referent \tabref{tab:2:findthereferent}\todo{maybe: see Table 2.1??}.

@@

I deliberately left an important aspect of language out. 
In the case of physical agents, the form cannot be transmitted
directly but needs to be articulated in speech sounds, 
written signs, or gestures, to create a true utterance. This additional 
complexity will not be discussed further in this book -- even though 
it is a fascinating topic in its own right.\footnote{
See for a state of the art review: \cite{Clark:1990}. 
We have already been conducting quite extensive 
research in our group on the origin of sound systems 
using similar principles as the ones discussed in this 
book for the origins of word semantics. 
See: \cite{DeBoer:1997}. The complex 
adaptive systems approach underlying this phonetics
work was already foreshadowed in work of phoneticians
like \cite{Liljencrants:1972}.}
To make a verbal interaction nevertheless complete, agents
are given a repertoire of consonants and vowels with 
which they can make random syllables and syllable combinations, 
like `wabido', `bimaku', etc. The articulation and 
recognition of these syllables is assumed to be acquired
already and transmission is engineered to be error-free. 
This way, our attention can be focused on how
ontologies, lexicons, and grammars may emerge.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap2/figs/hpos.pdf}}
\caption{Three examples of segmented images (from top to bottom). Left and right images 
show the perceived and segmented images for two agents respectively. These images are always
different because they are dependent on the position of the agent. 
The topic is indicated by a dashed bounding box in the 
image of the speaker (on the right for the first case and on the left for the others). 
Segments which are too small are ignored. The topics have all been conceptualised
as being to the right and so the same word 
`gofubo' has been used to successfully refer to them.}
\label{f:plate10}
\end{figure}

Segmentation can happen according to several criteria. \is{image segmentation}
For example, patches with similar colour can be grouped 
into a single patch, edges can be identified 
and then linked with each other to form line segments
circumscribing the contours of 
an object, or consecutive images can be compared to extract the parts
that changed and hence moved as a single object. 
There is now a solid body of techniques from decades
of machine vision research to efficiently segment scenes according to 
these and other methods. The Talking Heads use several methods
in parallel and combine their output to get a clearly segmented
picture.\footnote{
For general introductions to these areas, 
see: \cite{Ballard:1982} and \cite{Fischler:1987}.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap2/figs/scene1-1.pdf}}
\caption{\label{scene1-1}The scene contains two 
objects: a triangular object and a rectangular object
with the same gray levels. 
Each one is characterised by values on three sensory 
channels: HPOS, VPOS, and GRAY.}
\end{figure}

The low-level visual processes outputting
values on the various sensory channels are already quite
complicated in themselves. I will discuss them further in 
\chapref{chap:3}. I will then argue that the 
agent's repertoire of visual processes does not have to be
static and programmed in advance but evolves and adapts
through a selectionist process. New processes may ``grow''
by the random combination of primitive operations and are
pruned when they fail to yield useful information or information 
which is irrelevant in the environment in which the agent 
finds himself.

@@

The data on sensory channels are values from a continuous domain
(between 0.0 and 1.0). 
To be the basis of natural language communication, these
values must be transformed into a discrete domain. This is 
precisely the task of the conceptual layer. It 
performs massive data reduction to make infinitely
rich environments managable. One means of categorisation
is to divide up each domain of values on a particular
sensory channel into regions and 
simply assign a category to each region, thus creating a 
discrimination tree.\is{discrimination tree} For example, 
the HPOS-channel can be cut in two halves leading to a 
distinction between [LEFT] ($0.0 \leq HPOS < 0.5$) and [RIGHT] 
($0.5 \leq HPOS \leq 1.0$). The triangular object 
in \figref{scene1-1} has the value HPOS=0.35 
and would therefore be categorised as [LEFT]. Similarly, the 
VPOS-channel can be divided in two halves yielding the 
categories [LOWER] and [UPPER], and likewise the GRAY-channel
yielding the categories [LIGHT] and [DARK]. 
Given these categories, 
the rectangular object in the scene of 
\figref{scene1-1} would be categorised as [RIGHT UPPER LIGHT].
Of course, light, dark, left, lower, etc. are labels 
that I have given to these categories. The Talking Heads 
create categories by partitioning sensory channels but do
not use these labels internally.

It is always possible to refine a distinction by further
subdividing its region. Thus an agent could further divide the 
bottom region of the HPOS-channel (categorised as 
[LEFT]) into two subregions [TOTALLY-LEFT] ($0.0 \le HPOS < 0.25$), 
and [MID-LEFT] ($0.25 \le HPOS < 0.5$). The triangular object
can now be categorised as [MID-LEFT], if simply categorising it
as [LEFT] is not distinctive enough. Each of these categories can 
still be further refined. The categorisation networks
resulting from these consecutive binary subdivisions form
discrimination trees (\figref{tree1}). It is not at
all assumed that all agents have the same trees; due to 
different developmental histories, divergence is inevitable.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.35\textwidth]{chap2/figs/tree1.pdf}}
\caption{\label{tree1}A discrimination tree displays
the divisions of the total range of 
values on a sensory channel into finer and finer subregions. 
Categories are assigned to each region at different 
levels of the tree.}
\end{figure}

There are obviously other ways to 
move from the continuous domain of sensory channels to 
the discrete domain of categories.\footnote{
See: \cite{Taylor:1989}.}
For example, it is not really necessary 
to have a binary division, we could just as well split 
each region into three or more subregions. Or we could introduce
focal (prototypical) values and associate
a category with each of them. In the latter case, 
the categorisation process consists 
in identifying the prototype that is closest to 
an object's value. For the current experiment, I will stick however
to a binary categorisation strategy because that is
the simplest to understand and formally investigate.

We will see that agents build hundreds,
even thousands, of categories as they play their
language games, and in addition they make combinations of 
categories. To bring some order in this profusion of 
categories, I will label them using the 
sensory channel from which a category operates, 
followed by the upper and lower bound of the region they 
carve out. Thus [TOTALLY-LEFT]
is labeled as [HPOS 0.0–0.25], because it carves 
out a region between 0.0 and 0.25 on the HPOS-channel. 
When it must be emphasised that 
a category belongs to a particular agent, for example {\bfshape a1}, we
write [HPOS 0.0–0.25]$_{\mathbf a1}$. The same category in agent {\bfshape a2}
is labeled as [HPOS 0.0–0.25]$_{\mathbf a2}$. If I want to talk about 
this category in the abstract, I will not mention any 
agent and simply write [HPOS 0.0–0.25]. We will also 
allow conjunctive combinations of categories which will be
written as a set, as in \{[HPOS 0.0–0.25], [VPOS 0.0–0.25]\}, 
which could be paraphrased as totally-left {\itshape and}
totally-down.
